### 2016-03-29 PaperThoughts.Rnw
managed to wade my way through the entire process of ML estimation for SAOMs. I think I understand it fully now.

\documentclass[a4paper]{report}
\usepackage[margin=.75in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,array}
\usepackage{booktabs}
\usepackage{tikz}
%% load any required packages here

@@

\item \textbf{Probability of Change (nodes)} - Given that a change, call it $X(t) \equiv x \rightarrow X(t+1)$ occurs, each node has its own probability of changing. Recall that from time $t$ to time $t+1$, only one tie is changing (call it $x_ij$), so that only one node, $i$, is given the power to change one of its $n-1$ available ties from $x_ij$ to $1-x_{ij}$. The probability that node $i$ will be the one to change a tie, given $x \rightarrow X(t+1)$ is $\pi_i(\alpha, x) \equiv \pi_i(\alpha, x|x\rightarrow X(t+1)) = \frac{\lambda_i(\alpha, x)}{\lambda(\alpha, x)}$.

@@

\item \textbf{Moving from $x(t_{m-1})$ to $x(t_m)$} - First define $x(t_1)$ as the first network observed and $x(t_M)$ as the last network observed. Networks in between are written $x(t_{m})$ for $m = 2, \dots, M-1$. The observations $x(t_1), x(t_2), \dots, x(t_M)$ are embedded in the continuous-time Markov chain $\{X(t)|t_1 \leq t \leq t_M\}$. Moving from the observation $x(t_{m-1})$ to the observation $x(t_m)$ requires $C$ changes or steps in the Markov process $X(t)$ where $C = \sum_{i,j \in \mathcal{N}} |x_{ij}(t_m) - x_{ij}(t_{m-1})|$. Denote the $C-1$ timepoints in the the Markov process that correspond to the unobserved changes in the network as $T_1, T_2, \dots, T_{C-1}$ and define $T_0 = t_{m-1}$ and $T_C = t_{m}$. For each timepoint $T_c$, for $c = 1, \dots C-1$, there is a single actor, denoted $I_c$ that gets an opportunity to change at this timepoint. This actor changes one of its ties to one node, $J_c$. If there is no change, $I_c = J_c$. Formally, the pair $(I_c,J_c)$ is the only $(i,j)$ pair for which $x_{ij}(T_{c-1}) \neq x_{ij}(T_c)$ if such a pair exists, and $(I_c, J_c) = (I_c, I_c)$ otherwise. The triplet $(T_c, I_c, J_c)$ forms a stochastic process for $c = 1, \dots, C-1$ that, given the prior observation, $x(t_{m-1})$ for $m = 2, \dots, M$, completely determines $x(t)$ for $t_{m-1} < t < t_m$. 
\item \textbf{Augmenting Data} (see section on ML estimation) - The augmenting data consist of $C$ and the pairs $(I_c, J_c)$ for $c = 1, \dots, C-1$. (Note the timepoints $T_c$ are left out of the augmenting data.) The possible outcomes of the augmenting data are determined by the condition that if for a given tie, $x_{ij}(t_{m-1})$, the number of timepoints $1 \leq c \leq C-1$ at which $(i_c, j_c) = (i,j)$ is even if $x_{ij}(t_{m}) = x_{ij}(t_{m-1})$ and is odd if $x_{ij}(t_{m}) \neq x_{ij}(t_{m-1})$. 
\item \textbf{Sample Path} - The sample path is the stochastic process $V = ((I_c, J_c):c = 1, \dots, C-1)$. The elements of $V$ where $I_c = J_c$ are redundant. However, they are kept in the process because they help with computation of the likelihood. For notational convenience, define $x^{(c)} = x(T_C)$. Then, the networks $x^{(c)}$ and $x^{(c-1)}$ only differ in the element $(I_c, J_c) \in V$. 

@@

\section{Important Facts about SAOMs}
\begin{itemize}
\item In SAOMs, the observed networks in a longitudinal study, call them $x_{t_1}, \dots, x_{t_M}$, are discrete observations embedded in a continuous time process, $X(t)$. This means that the probability model is defined \emph{independently} of the observational design. Thus, the time between observations can be irregular and still be analyzed under this model.
\item The continuous-time Markov process $\{X(t) | t_1 \leq t \leq t_M\}$ has a distribution that consitutes an exponential family, which are of the form $f(y|\theta) = \exp\{\sum_{j=1}^s q_j(\theta) T_j(y) \} c(\theta) h(y)$ where $y$ is the data vector and $\theta$ is the scalar or vector parameter. This means the SAOMs can be viewed as ``incompletely observed exponential family" models. 
\item Inference is computer-intensive and time-consuming because of ``elaborate MCMC procedures" that must be implemented.
\item The definition of SAOMs means that data can be simulated directly from the probability distribution given the initial network state $x_{t_1}$. 
\item However, SAOMs are ``too complicated for the calculation of likelihoods or estimators in closed form." The distribution, however, can be easily simulated from, so that estimation can proceed by MCMC simulation.
\end{itemize}

\section{Maximum Likelihood Estimation}
Sketch of Algorithm:
\begin{enumerate}
\item Use the first network observation, $x(t_1)$, as a starting value. 
\item For $m = 2, 3, \dots, M$ where $M$ is the total number of observed networks, augment the observed data with random draws from the sequence of intermediate steps in the Markov process, $x(t)$, that could have led from $x(t_{m-1})$ to $x(t_m)$. i.e. draw from the set {\small $$\{(x(t^*_1), x(t^*_2), \dots x(t^*_{C-1}))' | x(t^*_1) \in \mathcal{A}_{1:n}(x(t_{m-1})) \& x(t^*_2) \in \mathcal{A}_{1:n}(x(t^*_1)) \& x(t^*_3) \in \mathcal{A}_{1:n}(x(t^*_2)) \& \cdots \& x(t_{m}) \in \mathcal{A}_{1:n}(x(t^*_{C-1})) \}$$ } where $C = \sum_{i,j \in \mathcal{N}} |x_{ij}(t_m) - x_{ij}(t_{m-1})|$. Note that $C$ in the number of single edge changes needed to get from $x(t_{m-1})$ to $x(t_m)$, so $C-1$ is the number of intermediate networks. These draws can be simulated using a Metropolis-Hastings algorithm. 
\item Use the draws from 2 as updates in a Robbins-Monro algorithm (?) to find the solution of the likelihood equation. 
\end{enumerate}

>>>>

\documentclass[a4paper]{report}
\usepackage[margin=.75in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{amsmath,amssymb,array}
\usepackage{booktabs}
\usepackage{tikz}
%% load any required packages here

@@

\item \textbf{Probability of Change (nodes)} - Given that a change, call it $X(t) \equiv x \rightarrow X(t+1)$ occurs, each node has its own probability of changing. Recall that from time $t$ to time $t+1$, only one tie is changing (call it $x_{ij}$), so that only one node, $i$, is given the power to change one of its $n-1$ available ties from $x_ij$ to $1-x_{ij}$. The probability that node $i$ will be the one to change a tie, given $x \rightarrow X(t+1)$ is $\pi_i(\alpha, x) \equiv \pi_i(\alpha, x|x\rightarrow X(t+1)) = \frac{\lambda_i(\alpha, x)}{\lambda(\alpha, x)}$.

@@

\item \textbf{Moving from $x(t_{m-1})$ to $x(t_m)$} - First define $x(t_1)$ as the first network observed and $x(t_M)$ as the last network observed. Networks in between are written $x(t_{m})$ for $m = 2, \dots, M-1$. The observations $x(t_1), x(t_2), \dots, x(t_M)$ are embedded in the continuous-time Markov chain $\{X(t)|t_1 \leq t \leq t_M\}$. Moving from the observation $x(t_{m-1})$ to the observation $x(t_m)$ requires $R$ changes or steps in the Markov process $X(t)$ where $R = \sum_{i,j \in \mathcal{N}} |x_{ij}(t_m) - x_{ij}(t_{m-1})|$. Note that the observed network change from $x(t_{m-1})$ to $x_(t_m)$ is conditionally independent of all other network observations given $x(t_{m-1})$. Denote the $R$ timepoints in the the Markov process that correspond to the unobserved changes in the network as $T_1, T_2, \dots, T_{R-1}$ and define $T_0 = t_{m-1}$ and $T_R \leq t_{m} < T_{R+1}$. For each timepoint $T_r$, for $r = 1, \dots R$, there is a single actor, denoted $I_r$ that gets an opportunity to change at this timepoint. This actor changes one of its ties to one node, $J_r$. If there is no change, $I_r = J_r$. Formally, the pair $(I_r,J_r)$ is the only $(i,j)$ pair for which $x_{ij}(T_{r-1}) \neq x_{ij}(T_r)$ if such a pair exists, and $(I_r, J_r) = (I_r, I_r)$ otherwise. The triplet $(T_r, I_r, J_r)$ forms a stochastic process for $r = 1, \dots, R$ that, given the prior observation, $x(t_{m-1})$ for $m = 2, \dots, M$, completely determines $x(t)$ for $t_{m-1} < t < t_m$. 
\item \textbf{Augmenting Data} (see section on ML estimation) - The augmenting data consist of $R$ and the pairs $(I_r, J_r)$ for $r= 1, \dots, R$. (Note the timepoints $T_r$ are left out of the augmenting data.) The possible outcomes of the augmenting data are determined by the condition that if for a given tie, $x_{ij}(t_{m-1})$, the number of timepoints $1 \leq r\leq R$ at which $(i_r, j_r) = (i,j)$ is even if $x_{ij}(t_{m}) = x_{ij}(t_{m-1})$ and is odd if $x_{ij}(t_{m}) \neq x_{ij}(t_{m-1})$. 
\item \textbf{Sample Path} - The sample path is the stochastic process $V = ((I_r, J_r):r = 1, \dots, R)$. Note that the elements of $V$ where $I_r = J_r$ are redundant. However, they are kept in the process because they help with computation of the likelihood. For notational convenience, define $x^{(r)} = x(t_r)$. Using this notation, write $x(t_{m-1}) = x^(0)$ for the independent cases where $m = 2, \dots, M$. Then, the networks $x^{(r)}$ and $x^{(r-1)}$ only differ in the element $(I_r, J_r) \in V$, provided that $I_r\neq J_r$.
\item \textbf{PDF of Sample Path} - The sample path has probability distribution function 
  $$ p_{sp}(V|\alpha, \beta, x^{(0)} ) = Pr(T_R \leq t_m < T_{R+1} | x^{(0)}, V, \alpha, \beta) \times \prod_{r=1}^R \pi_{i_r}(\alpha, x^{(r-1)})p_{i_r,j_r}(\beta,x^{(r-1)}). $$
The first component, $Pr(T_R \leq t_m < T_{R+1} | x^{(0)}, V, \alpha, \beta)$ is the probability that your next network observation at time point $t_m$ comes \emph{before} the next change in the continuous time Markov chain, $X(T_{R+1})$. From a discussion that follows, $Pr(T_R \leq t_m < T_{R+1} | x^{(0)}, V, \alpha, \beta)$, the probability that your next network observation at time point $t_m$ comes \emph{before} the next change in the continuous time Markov chain, $X(T_{R+1})$, can be approximated by 
$$ \kappa(\alpha, x(t_{m-1}), V) \approx \int_{t_{m-1}}^{t_m} p_{T_R}(s) \text{Pr}(\{T_{R+1} - T_R > t_{m} - T_R \} | T_R = s) ds $$
$T_{R+1}$ has distribution Exponential$(\lambda(\alpha, x^{(R)}))$. So, Pr$(\{T_{R+1} - T_R > t_m - T_R \}) = \text{Pr}(T_{R+1} > t_m) = 1 - \text{Pr}(T_{R+1} < t_m) = 1 - (1 - \exp\{-\lambda(\alpha, x^{(R)})t_m\}) = \exp\{-\lambda(\alpha, x^{(R)})t_m\}$. So, the above integral can be written as   
$$ \kappa(\alpha, x(t_{m-1}), V) \approx \int_{t_{m-1}}^{t_m} p_{T_R}(s) \exp\{-\lambda(\alpha, x^{(R)})(t_m-s)\}. ds $$
This quantity is approximately equal to $\frac{p_{T_R}(t_m)}{\lambda(\alpha, x^{(R)})}$. The second expression, $\prod_{r=1}^R \pi_{i_r}(\alpha, x^{(r-1)})p_{i_r,j_r}(\beta,x^{(r-1)})$, is the PDF of the stochastic process $V$. So, the probability of the sample path can be approximated by:
$$ p_{sp}(V|\alpha, \beta, x^{(0)}) \approx \frac{p_{T_R}(t_m)}{\lambda(\alpha, x^{(R)})} \times \prod_{r=1}^R \pi_{i_r}(\alpha, x^{(r-1)})p_{i_r,j_r}(\beta,x^{(r-1)}). $$
So, for observed data, augmented by the sample path, ``the likelihood conditional on $x(t_{m-1})$ can be expressed directly, either exactly [(for constant $\lambda_i$)] or in good approximation."
\item \textbf{Distribution of waiting times between $x(t_{m-1})$ and $x(t_m)$} - Let $T_{r+1} - T_r$ be the waiting time between one of the $r = 1, \dots, R$ changes between $x(t_{m-1})$ and $x(t_m)$. After conditioning on $x(t_{m-1}) \equiv x^{(0)}$ and $V$, these waiting times have distribution $T_{r+1} - T_{r} \sim \text{Exponential}(\lambda(\alpha, x^{(r)}))$ for $r = 0, 1, \dots, R-1$. Following from this, the distribution of $T_R - t_{m-1}$, the time it takes for the network to move from state $x(t_{m-1})$ to state $x(t_{m})$ is the convolutions of the Exponential$(\lambda(\alpha, x^{(r)}))$ distributions for $r = 0, \dots, R-1$. The pdf of $T_R - t_{m-1}$ is\footnote{Source: \url{http://www.ccms.or.kr/data/pdfpaper/jcms21_4/21_4_501.pdf}}: 
$$ h(T_R - t_1 | \alpha, x(t_{m-1}), V) = \sum_{r=0}^{R-1} \frac{\lambda(\alpha, x^{(0)})\cdot \lambda(\alpha, x^{(1)}) \cdot \lambda(\alpha, x^{(R-1)})}{\prod_{q\neq r, q = 0}^{R-1} (\lambda(\alpha, x^{(q)}) - \lambda(\alpha, x^{(r)}))} \cdot \exp\{-(T_R - t_{m-1})\cdot \lambda(\alpha, x^{(r)})\}$$
In the special case that the $\lambda_{i}(\alpha, x)$ rates are constant, call it $\alpha_1$, then the number of steps in the Markov chain to move from $x(t_{m-1})$ to $x(t_m)$, $R$ has distribution $R \sim \text{Poisson}(n\alpha_1(t_{m} - t_{m-1}))$. In this case, $Pr(T_R \leq t_m < T_{R+1} | x^{(0)}, V, \alpha, \beta) = \kappa(\alpha_1, x^{(0)}, V) = \exp\{-n\alpha_1(t_{m} - t_{m-1})\}\frac{(n\alpha_1(t_m - t_{m-1}))^R}{R!}$. However, when the individual rates are different for each actor, this probability must be approximated. 
\item \textbf{Distribution of $T_R$} - $T_R$ is the timepoint in the Markov chain at which the state $x(t_m)$ is acheived. Since $T_R - t_{m-1}$ is a convolution of exponentials with rate parameters $\lambda(\alpha, x^{(r)})$ for $r = 0, R-1$, by the Lyapounov Central Limit Theorem, the distribution of $T_R$ is approximately normal with mean $\mu_{\alpha}$ and variance $\sigma^2_{\alpha}$ where 
$$ \mu_{\alpha} = \sum_{r=0}^{R-1} \frac{1}{\lambda(\alpha, x^{(r)})} \quad \text{and} \quad  \sigma^2_{\alpha} = \sum_{r=0}^{R-1} \frac{1}{\lambda(\alpha, x^{(r)})^2} $$
The PDF of $T_R$ is thus $p_{T_R} \approx \frac{1}{\sqrt{2\pi \sigma^2_{\alpha}}} \exp\left\{\frac{-((t-t_{m-1})-\mu_{\alpha})^2}{2\sigma^2_{\alpha}}\right\}$. 

@@

\section{Important Facts about SAOMs}
\begin{itemize}
\item In SAOMs, the observed networks in a longitudinal study, call them $x_{t_1}, \dots, x_{t_M}$, are discrete observations embedded in a continuous time process, $X(t)$. This means that the probability model is defined \emph{independently} of the observational design. Thus, the time between observations can be irregular and still be analyzed under this model.
\item The continuous-time Markov process $\{X(t) | t_1 \leq t \leq t_M\}$ has a distribution that consitutes an exponential family, which are of the form $f(y|\theta) = \exp\{\sum_{j=1}^s q_j(\theta) T_j(y) \} c(\theta) h(y)$ where $y$ is the data vector and $\theta$ is the scalar or vector parameter. This means the SAOMs can be viewed as ``incompletely observed exponential family" models. 
\item Inference is computer-intensive and time-consuming because of ``elaborate MCMC procedures" that must be implemented.
\item The definition of SAOMs means that data can be simulated directly from the probability distribution given the initial network state $x_{t_1}$. 
\item However, SAOMs are ``too complicated for the calculation of likelihoods or estimators in closed form." The distribution, however, can be easily simulated from, so that estimation can proceed by MCMC simulation.
\item ML estimation of SAOMs is based on the missing data principle. The ``missing data" here are the paths from $x(t_{m-1})$ to $x(t_{m})$ for all $m = 2, \dots, M$. This set can be denoted for each $m$ by $V_m = \{(I_{m1}, J_{m1}),\dots, (I_{mR_m}, J_{mR_m})\}$. So, the complete set of augmenting data is $V = (V_2, \dots V_M)$.  
\end{itemize}

\section{Maximum Likelihood Estimation}
Sketch of Algorithm:
\begin{enumerate}
\item Use the first network observation, $x(t_1)$, as a starting value. All analysis proceeds conditioning on $x(t_1)$.  
\item For $m = 2, 3, \dots, M$ where $M$ is the total number of observed networks, augment the observed data with random draws from the sequence of intermediate steps in the Markov process, $x(t)$, that could have led from $x(t_{m-1})$ to $x(t_m)$. i.e. draw from the set {\small $$\{(x(t^*_1), x(t^*_2), \dots x(t^*_{C-1}))' | x(t^*_1) \in \mathcal{A}_{1:n}(x(t_{m-1})) \& x(t^*_2) \in \mathcal{A}_{1:n}(x(t^*_1)) \& x(t^*_3) \in \mathcal{A}_{1:n}(x(t^*_2)) \& \cdots \& x(t_{m}) \in \mathcal{A}_{1:n}(x(t^*_{C-1})) \}$$ } where $C = \sum_{i,j \in \mathcal{N}} |x_{ij}(t_m) - x_{ij}(t_{m-1})|$. Note that $C$ in the number of single edge changes needed to get from $x(t_{m-1})$ to $x(t_m)$, so $C-1$ is the number of intermediate networks. These draws can be simulated using a Metropolis-Hastings algorithm. 
\item Use the draws from 2 as updates in a Robbins-Monro algorithm (?) to find the solution of the likelihood equation. 
\end{enumerate}

Full Iterative Algorithm: For each iteration $N = 1, 2, \dots$
\begin{enumerate}
\item For each $m = 2, \dots, M$, make a large number of Metropolis-Hastings steps of the following form.  Let $\underline{v} = ((i_1, j_1), \dots, (i_R, i_R))$ be a given path from $x(t_{m-1})$ to $x(t_m)$ in the whole possible set of paths $\mathcal{V}$. Then propose a new path, $\tilde{\underline{v}}$ from the proposal distribution that consists of all of the possible small changes: 
  \begin{enumerate}
  \item ``Paired Deletions" - Of all pairs of indices $r_1, r_2$ such that $(i_{r_1}, j_{r_1}) = (i_{r_2}, j_{r_2})$, $i_{r_1} \neq j_{r_1}$, randomly select a pair $(r_1, r_2)$ and delete both from the current path.
  \item ``Paired Insertions" - Randomly select an edge $(i,j) \in \mathcal{N}^2$ with $i \neq j$. Randomly choose 2 indices, $r_1, r_2$. Insert $(i,j)$ immediately before $r_1$ and $r_2$.  
  \item ``Single Insertions" - At a random place in the current path, insert $(i,i)$ for a random $i \in \mathcal{N}$.
  \item ``Single Deletions" - Of all elements in the current path that satisfy $i_r = j_r$, randomly delete one of them. 
  \item ``Permutations" - For randomly chosen $r_1 < r_2$, where $r_2 - r_1$ is bounded from above by some smallish number to avoid lengthy computation (tuning parameter?), the segment of consecutive elements $(i_r,j_r)$, $r = r_1, \dots, r_2$ is randomly permuted. 
  \end{enumerate}
Denote the proposal probabilities $u(\tilde{\underline{v}} | \underline{v})$ and the target probabilities $p(\underline{v})$. The target distribution is $p(\underline{v}) \propto p_{SP}(\underline{v} | \alpha, \beta, x(t_{m-1}))$. Then the acceptance probability for a proposal path $\tilde{\underline{v}}$ and a current path $\underline{v}$ is: $\min \left\{ 1, \frac{p(\tilde{\underline{v}})u(\underline{v}|\tilde{\underline{v}})}{p(\underline{v})u(\tilde{\underline{v}}| \underline{v})}\right\}$. The series of M-H steps will result in a new set of augmenting data, $(v^{(N)}) = (v_2^{(N)}, \dots, v_M^{(N)})$. 
\item Compute 
  $$ S_{XV}(\hat{\theta^{(N)}}; x, v^{(N)}) = \sum_{m=2}^M S_m(\hat{\theta^{(N)}}; x(t_{m-1}), v_m^{(N)}) $$
$S_{XV}(\hat{\theta^{(N)}}; x, v^{(N)})$ is the complete data score function. $S_m(\hat{\theta^{(N)}}; x(t_{m-1}), v_m^{(N)}) = \frac{\partial \log p_m(v_m; \theta | x(t_{m-1}))}{d\theta}$ is the total data score function at step $m$. $p_m$ is $p_{SP}(v_m; \theta = (\alpha, \beta)' | x(t_{m-1}))$, the probability of the sample path $v_m$. 
\item Update $$\theta^{(N+1)} =(\alpha^{(N+1)}, \beta^{(N+1)}) = \theta^{(N)} + a_N D^{-1} S_{XV}(\hat{\theta^{(N)}}; x, v^{(N)})$$
where $D$ is a MC estimate of the complete data observed Fisher information matrix, $D_{XV}(\theta) = -\frac{\partial S_{XV}(\theta: x, V)}{\partial \theta}$ estimated for $\theta = \theta^{(1)}$. (i.e. evaluated at the starting value of $\theta$.) The starting value is the method of moments estimator, $\hat{\theta}^{(1)}$. (See Snijders 2001 for how to get the MoM estimate). The sequence $a_N$ is just a sequence that approaches 0 as $N \to \infty$. (I'm not sure exactly what? Maybe just $\frac{1}{N}$ would work? There's no specifics in Snijders 2001, and I'm looking through the RSiena manual for clues...). 
\end{enumerate}
